Appendix 1: Evaluation of the three species in Figure 1

### 2015-08-05 Harris_2015/manuscript-materials/draft.tex
submitted to Ecology

\textbf{Abstract:} Estimating species interactions from observational
data is one of the most controversial tasks in community ecology. One
difficulty is that a single pairwise interaction can ripple through an
ecological network and produce surprising indirect consequences. For
example, two competing species would ordinarily correlate negatively in
space, but this effect can be reversed in the presence of a third
species that is capable of outcompeting both of them when it is present.
Here, I apply models from statistical physics, called Markov networks or
Markov random fields, that can predict the direct and indirect
consequences of any possible species interaction matrix. Interactions in
these models can be estimated from observational data via maximum
likelihood. Using simulated landscapes with known pairwise interaction
strengths, I evaluated Markov networks and several existing approaches.
The Markov networks consistently outperformed other methods, correctly
isolating direct interactions between species pairs even when indirect
interactions or abiotic environmental effects largely overpowered them.
Two computationally efficient approximations, based on linear and
generalized linear models, also performed well. Indirect effects
reliably caused a common null modeling approach to produce incorrect
inferences, however.

@@

\setlength{\parindent}{1em} \setlength{\parskip}{3pt}

If nontrophic species interactions, such as competition, are important
drivers of community assembly, then ecologists might expect to see their
influence in species composition data (MacArthur 1958, Diamond 1975).
Despite decades of work and several major controversies, however (Lewin
1983, Strong et al. 1984, Gotelli and Entsminger 2003, Connor et al.
2013), existing methods for detecting competition's effects on community
structure are unreliable (Gotelli and Ulrich 2009). In particular,
species' effects on one another can become lost in the complex web of
direct and indirect interactions in real assemblages. For example, the
competitive interaction between the two shrub species in Figure 1A can
become obscured by their shared tendency to occur in unshaded areas
(Figure 1B). While ecologists have long known that indirect effects can
overwhelm direct ones at the landscape level (Dodson 1970, Levine 1976),
the vast majority of our methods for drawing inferenes from
observational data do not control for these effects (e.g. Diamond 1975,
Strong et al. 1984, Gotelli and Ulrich 2009, Veech 2013, Pollock et al.
2014). To the extent that indirect interactions like those in Figure 1
are generally important (Dodson 1970), existing methods will thus not
generally provide much evidence regarding species' direct effects on one
another. The goal of this paper is to resolve this long-standing
problem.

While competition doesn't reliably leave the expected pattern in
community structure (low co-occurrence rates at the landscape level), it
nevertheless does leave a signal in the data (Figure 1C). Specifically,
\emph{among shaded sites}, there will be a deficit of co-occurrences
between the competing shrub species, and \emph{among unshaded sites},
there will also be such a deficit. More generally, we can obtain much
better estimates of the association between two species from their
conditional relationships (i.e.~by controlling for other species in the
network) than we could get from their overall co-occurrence rates. This
kind of precision is difficult to obtain from null models, which begin
with the assumptions that all the pairwise interactions are zero and
thus don't need to be controlled for. Nevertheless, null models have
dominated this field for more than three decades (Strong et al. 1984,
Gotelli and Ulrich 2009).

Following recent work by Azaele et al. (2010) and Fort (2013), this
paper shows that Markov networks (undirected graphical models also known
as Markov random fields; Murphy 2012) can provide a framework for
understanding the landscape-level consequences of pairwise species
interactions, and for detecting them from observed presence-absence
matrices. Markov networks have been used in many scientific fields for
decades in similar contexts, from physics (where nearby particles
interact magnetically; Cipra 1987) to spatial statistics (where adjacent
grid cells have correlated values; Harris 1974, Gelfand et al. 2005).
While community ecologists explored some related approaches in the
1980's (Whittam and Siegel-Causey 1981), they used severe approximations
that led to unintelligible results (e.g. ``probabilities'' greater than
one; Gilpin and Diamond 1982).

@@

Markov networks provide a framework for translating back and forth
between the conditional relationships among species (Figure 1C) and the
kinds of species assemblages that these relationships produce. Here, I
show how a set of conditional relationships can be used to determine the
how groups of species can co-occur. Methods for estimating conditional
relationships from data are discussed in the next section.

@@

In the previous section, the values of \(\alpha\) and \(\beta\) were
known and the goal was to make predictions about possible species
assemblages. In practice, however, ecologists will often need to
estimate the parameters from an observed co-occurrence matrix (i.e.~from
a matrix of ones and zeros indicating which species are present at which
sites). When the number of species is reasonably small, one can compute
exact maximum likelihood estimates for all of the \(\alpha\) and
\(\beta\) coefficients given a presence-absence matrix by optimizing
\(p(\vec{y}; \alpha, \beta)\). Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces (Murphy 2012),
ensuring that this procedure will always converge on the global maximum.
This maximum represents the unique combination of \(\alpha\) and
\(\beta\) coefficients that would be expected to produce exactly the
observed co-occurrence frequencies on average (i.e.~it matches the
sufficient statistics of the model distribution to the sufficient
statistics of the data; Murphy 2012). I used the rosalia package (Harris
2015a) for the R programming language (R Core Team 2015) to optimize the
Markov network parameters. The package was named after Santa Rosalia,
the patron saint of biodiversity, whose supposedly miraculous healing
powers played an important rhetorical role in the null model debates of
the 1970's and 1980's (Lewin 1983).

@@

I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).

First, I used the rosalia package (Harris 2015a) to fit Markov newtork
models, as described above. For the analyses with 20 species, I added a
very weak logistic prior distribution on the \(\alpha\) and \(\beta\)
terms with scale 2 to ensure that the model estimates were always
finite. The bias introduced by this prior should be small: the 95\%
credible interval on \(\beta\) only requires that one species' effect on
the odds of observing a different species to be less than a factor of
1500 (which is not much of a constraint). The logistic distribution was
chosen because it is convex and has a similar shape to the Laplace
distribution used in LASSO regularization (especially in the tails), but
unlike the Laplace distribution it is differentiable everywhere and does
not force any estimates to be exactly zero. To confirm that this
procedure produced stable estimates, I compared its estimates on 50
bootstrap replicates (Appendix 4).

I also evaluated five alternative methods from the existing literature
(plus a novel variant of the fifth method). The first two alternative
interaction measures were the sample correlations and the partial
correlations between each pair of species' data vectors on the landscape
(Albrecht and Gotelli 2001, Faisal et al. 2010). In the context of
non-Gaussian data, the partial correlation can be thought of as a
computationally efficient approximation to the full Markov network model
(Loh and Wainwright 2013). This sort of model is very common for
estimating relationships among genes and gene products (Friedman et al.
2008). Because partial correlations are undefined for landscapes with
perfectly-correlated species pairs, I used a regularized estimate based
on James-Stein shrinkage, as implemented in the corpcor package's
\texttt{pcor.shrink} function with the default settings (Schäfer et al.
2014).

@@

The fourth alternative method, described in Gotelli and Ulrich (2009),
involved simulating new landscapes from a null model that retains the
row and column sums of the original matrix (Strong et al. 1984). I used
the \(Z\)-scores computed by the Pairs software described in Gotelli and
Ulrich (2009) as my null model-based estimator of species interactions.

@@

I then evaluated the relationship between each method's estimates and
the ``true'' interaction strengths among all of the species pairs from
the larger simulated landscapes. This determined which the methods
provide a consistent way to know how strong species interactions
are---regardless of which species were present in a particular data set
or how many observations were taken. Because the different methods
mostly describe species interactions on different scales
(e.g.~correlations versus \(Z\) scores versus regression coefficients),
I used linear regression through the origin to rescale the different
estimates produced by each method so that they had a consistent
interpretation. After rescaling each method's estimates, I calculated
squared errors between the scaled interaction estimates and ``true''
interaction values across all the simulated data sets. These squared
errors determined the proportion of variance explained for different
combinations of method and landscape size (compared with a null model
that assumed all interaction strengths to be zero).

@@

Bootstrap resampling indicated that the above ranking of the different
methods was robust (Appendix 3). In particular, the 95\% confidence
interval of the bootstrap distribution indicated that the Markov
network's overall \(R^2\) value was between 14 and 18 percent higher
than the second-most effective method (generalized linear models) and
between 2.12 and 2.38 times higher than could be achieved by the null
model (Pairs). Bootstrap resampling of a 200-site landscape also
confirmed that the rosalia package's estimates of species' conditional
relationships were reasonably robust to sampling variation (Appendix 4).

@@

Despite their limitations, Markov networks have enormous potential to
improve ecological understanding. In particular, they are less
vulnerable than some of the most commonly-used methods to mistakenly
identifying positive species interactions between competing species, and
can make precise statements about the conditions whre indirect
interactions will overwhelm direct ones. They also provide a simple
answer to the question of how competition should affect a species'
overall prevalence, which was a major flashpoint for the null model
debates in the 1980's (Roughgarden 1983, Strong et al. 1984). Equation 1
can be used to calculate the expected prevalence of a species in the
absence of biotic influences (\(\frac{1}{1 + e^{-\alpha}}\); Lee and
Hastie 2012). Competition's effect on prevalence in a Markov network can
then be calculated by subtracting this value from the observed
prevalence (cf Figure 2D). This kind of insight would have been
difficult to obtain without a generative model that makes predictions
about the consequences of species interactions; null models (which
presume \emph{a priori} that interactions do not exist) have no way to
make such predictions.

Markov networks---particularly the Ising model for binary
networks---have been studied in statistical physics for nearly a century
(Cipra 1987), and the models' properties, capabilities, and limits are
well-understood in a huge range of applications. Using the same
framework for species interactions would thus allow ecologists to tap
into an enormous set of existing discoveries and techniques for dealing
with indirect effects, stability, and alternative stable states.
Numerous other extensions are possible: for example, the states of the
interaction network can be modeled as a function of the local abiotic
environment (Lee and Hastie 2012), which would provide a rigorous and
straightforward approach to the difficult and important task of
incorporating whole networks of biotic interactions into species
distribution models (Kissling et al. 2012, Pollock et al. 2014), leading
to a better understanding of the interplay between biotic and abiotic
effects on community structure. There are even methods (Whittam and
Siegel-Causey 1981, Tjelmeland and Besag 1998) that would allow one
species to affect the sign or strength of the relationship between two
other species, tipping the balance between facilitation and exploitation
(Bruno et al. 2003).

Finally, the results presented here have important implications for
ecologists' continued use of null models for studying species
interactions. Null and neutral models can be useful for clarifying our
thinking about the numerical consequences of species' richness and
abundance patterns (Harris et al. 2011, Xiao et al. 2015), but
deviations from a particular null model must be interpreted with care
(Roughgarden 1983). Even in small networks with three species, it may
simply not be possible to implicate individual species pairs or specific
ecological processes like competition by rejecting a general-purpose
null (Gotelli and Ulrich 2009), especially when the test statistic is
effectively just a correlation coefficient (Figure 3B). Simultaneous
estimation of multiple ecological parameters seems like a much more
promising approach: to the extent that the models' relative performance
on real data sets is similar to the range of results shown in Figure 3A,
scientists in this field could often double their explanatory power by
switching from null models to Markov networks (or increase it
substantially by with linear or generalized linear approximations).
Regardless of the methods ecologists ultimately choose, controlling for
indirect effects could clearly improve our understanding of species'
direct effects on one another and on community structure.

@@

\textbf{Figure 3.} \textbf{A.} Proportion of variance in interaction
coefficients explained by each versus number of sampled locations.
\textbf{B.} The \(Z\)-scores produced by the null model (``Pairs'') for
each pair of species are very predictable from the correlation between
the presence-absence vectors of those same species and from the number
of sites on the landscape.

@@

Xiao, X., D. J. McGlinn, and E. P. White. 2015. A strong test of the
Maximum Entropy Theory of Ecology. The American Naturalist 185:E70--E80.
\end{flushleft}
\end{spacing}

>>>>

\textbf{Abstract:} Estimating species interactions from observational
data is one of the most controversial tasks in community ecology. One
difficulty is that a single pairwise interaction can ripple through an
ecological network and produce surprising indirect consequences. For
example, the negative correlation between two competing species can be
reversed in the presence of a third species that is capable of
outcompeting both of them. Here, I apply models from statistical
physics, called Markov networks or Markov random fields, that can
predict the direct and indirect consequences of any possible species
interaction matrix. Interactions in these models can be estimated from
observational data via maximum likelihood. Using simulated landscapes
with known pairwise interaction strengths, I evaluated Markov networks
and six existing approaches. The Markov networks consistently
outperformed other methods, correctly isolating direct interactions
between species pairs even when indirect interactions largely
overpowered them. Two computationally efficient approximations, based on
linear and generalized linear models, also performed well. Indirect
effects reliably caused a common null modeling approach to produce
incorrect inferences, however.

@@

\setlength{\parindent}{1cm}

To the extent that nontrophic species interactions (such as competition)
affect community assembly, ecologists might expect to find signatures of
these interactions in species composition data (MacArthur 1958, Diamond
1975). Despite decades of work and several major controversies, however
(Lewin 1983, Strong et al. 1984, Gotelli and Entsminger 2003, Connor et
al. 2013), existing methods for detecting competition's effects on
community structure are unreliable (Gotelli and Ulrich 2009). In
particular, species' effects on one another can become lost in the
complex web of direct and indirect interactions in real assemblages. For
example, the competitive interaction between the two shrub species in
Figure 1A can become obscured by their shared tendency to occur in
unshaded areas (Figure 1B). While ecologists have long known that
indirect effects can overwhelm direct ones at the landscape level
(Dodson 1970, Levine 1976), the vast majority of our methods for drawing
inferenes from observational data do not control for these effects (e.g.
Diamond 1975, Strong et al. 1984, Gotelli and Ulrich 2009, Veech 2013,
Pollock et al. 2014). To the extent that indirect interactions like
those in Figure 1 are generally important (Dodson 1970), existing
methods will thus not generally provide much evidence regarding species'
direct effects on one another. The goal of this paper is to resolve this
long-standing problem.

While competition doesn't reliably reduce co-occurrence rates at the
whole-landscape level (as most of our methods assume), it nevertheless
does leave a signal in the data (Figure 1C). Specifically, after
partitioning the data set into shaded sites and unshaded sites, there
will be co-occurrence deficits in each subset that might not be apparent
at the landscape level. More generally, we can obtain much better
estimates of the association between two species from their conditional
relationships (i.e.~by controlling for other species in the network)
than we could get from their overall co-occurrence rates. This kind of
precision is difficult to obtain from null models, which begin with the
assumption that all the pairwise interactions are zero and thus don't
need to be controlled for. Nevertheless, null models have dominated this
field for more than three decades (Strong et al. 1984, Gotelli and
Ulrich 2009).

Following recent work by Azaele et al. (2010) and Fort (2013), this
paper shows that Markov networks (undirected graphical models also known
as Markov random fields; Murphy 2012) can provide a framework for
understanding the landscape-level consequences of pairwise species
interactions, and for detecting them from observed presence-absence
matrices. Markov networks have been used in many scientific fields in
similar contexts for decades, from physics (where nearby particles
interact magnetically; Cipra 1987) to spatial statistics (where adjacent
grid cells have correlated values; Harris 1974, Gelfand et al. 2005).
While community ecologists explored some related approaches in the
1980's (Whittam and Siegel-Causey 1981), they used severe approximations
that led to unintelligible results (e.g. ``probabilities'' greater than
one; Gilpin and Diamond 1982).

@@

Markov networks provide a framework for translating back and forth
between the conditional relationships among species (Figure 1C) and the
kinds of species assemblages that these relationships produce. Here, I
show how a set of conditional relationships can be used to determine how
groups of species can co-occur. Methods for estimating conditional
relationships from data are discussed in the next section.

@@

In the previous section, the values of \(\alpha\) and \(\beta\) were
known and the goal was to make predictions about possible species
assemblages. In practice, however, ecologists will often need to
estimate the parameters from an observed co-occurrence matrix (i.e.~from
a matrix of ones and zeros indicating which species are present at which
sites). When the number of species is reasonably small, one can compute
exact maximum likelihood estimates for all of the \(\alpha\) and
\(\beta\) coefficients given a presence-absence matrix by optimizing
\(p(\vec{y}; \alpha, \beta)\). Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces (Murphy 2012),
ensuring that this procedure will always converge on the global maximum.
This maximum represents the unique combination of \(\alpha\) and
\(\beta\) coefficients that would be expected to produce exactly the
observed co-occurrence frequencies on average (i.e.~maximizing the
likelihood matches the sufficient statistics of the model distribution
to the sufficient statistics of the data; Murphy 2012). I used the
rosalia package (Harris 2015a) for the R programming language (R Core
Team 2015) to optimize the Markov network parameters. The package was
named after Santa Rosalia, the patron saint of biodiversity, whose
supposedly miraculous healing powers played an important rhetorical role
in the null model debates of the 1970's and 1980's (Lewin 1983).

@@

I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).
First, I used the rosalia package (Harris 2015a) to fit Markov newtork
models, as described above. For the analyses with 20 species, I added a
very weak logistic prior distribution on the \(\alpha\) and \(\beta\)
terms with scale 2 to ensure that the model estimates were always
finite. The bias introduced by this prior should be small: the 95\%
credible interval on \(\beta\) only requires that one species' effect on
the odds of observing a different species to be less than a factor of
1500 (which is not much of a constraint). The logistic distribution was
chosen because it is convex and has a similar shape to the Laplace
distribution used in LASSO regularization (especially in the tails), but
unlike the Laplace distribution it is differentiable everywhere and does
not force any estimates to be exactly zero. To confirm that this
procedure produced stable estimates, I compared its estimates on 50
bootstrap replicates (Appendix 4).

I also evaluated six alternative methods: five from the existing
literature, plus a novel combination of two of these methods. The first
alternative interaction metric was the sample correlation between
species' presence-absence vectors, which summarizes their marginal
association. Next, I used partial correlations, which summarize species'
conditional relationships (Albrecht and Gotelli 2001, Faisal et al.
2010). In the context of non-Gaussian data, the partial correlation can
be thought of as a computationally efficient approximation to the full
Markov network model (Loh and Wainwright 2013). This sort of model is
very common for estimating relationships among genes and gene products
(Friedman et al. 2008). Because partial correlations are undefined for
landscapes with perfectly-correlated species pairs, I used a regularized
estimate based on James-Stein shrinkage, as implemented in the corpcor
package's \texttt{pcor.shrink} function with the default settings
(Schäfer et al. 2014).

@@

The next method, described in Gotelli and Ulrich (2009), involved
simulating new landscapes from a null model that retains the row and
column sums of the original matrix (Strong et al. 1984). I used the
\(Z\)-scores computed by the Pairs software described in Gotelli and
Ulrich (2009) as my null model-based estimator of species interactions.

@@

I then evaluated the relationship between each method's estimates and
the ``true'' interaction strengths among all of the species pairs from
the larger simulated landscapes. This determined which of the methods
provide a consistent way to know how strong species interactions
are---regardless of which species were present in a particular data set
or how many observations were taken. Because the different methods
mostly describe species interactions on different scales
(e.g.~correlations versus \(Z\) scores versus regression coefficients),
I used linear regression through the origin to rescale the different
estimates produced by each method so that they had a consistent
interpretation. After rescaling each method's estimates, I calculated
squared errors between the scaled interaction estimates and ``true''
interaction values across all the simulated data sets. These squared
errors determined the proportion of variance explained for different
combinations of model type and landscape size (compared with a null
model that assumed all interaction strengths to be zero).

@@

Bootstrap resampling indicated that the above ranking of the different
methods was robust (Appendix 3). In particular, the 95\% confidence
interval of the bootstrap distribution indicated that the Markov
network's overall \(R^2\) value was between 14 and 18 percent higher
than the second-most effective method (generalized linear models) and
between 2.12 and 2.38 times higher than could be achieved by the null
model (Pairs). Bootstrap resampling of a 200-site landscape also
confirmed that the rosalia package's estimates of species' conditional
relationships were robust to sampling variation for reasonably-sized
landscapes (Appendix 4).

@@

Despite their limitations, Markov networks have enormous potential to
improve ecological understanding. In particular, they are less
vulnerable than some of the most commonly-used methods to mistakenly
identifying positive species interactions between competing species, and
can make precise statements about the conditions where indirect
interactions will overwhelm direct ones. They also provide a simple
answer to the question of how competition should affect a species'
overall prevalence, which was a major flashpoint for the null model
debates in the 1980's (Roughgarden 1983, Strong et al. 1984). Equation 1
can be used to calculate the expected prevalence of a species in the
absence of biotic influences (\(\frac{e^\alpha}{1 + e^{\alpha}}\); Lee
and Hastie 2012). Competition's effect on prevalence in a Markov network
can then be calculated by subtracting this value from the observed
prevalence (cf Figure 2D). This kind of insight would have been
difficult to obtain without a generative model that makes predictions
about the consequences of species interactions; null models (which
presume \emph{a priori} that interactions do not exist) have no way to
make such predictions.

Markov networks---particularly the Ising model for binary
networks---have been studied for nearly a century (Cipra 1987), and the
models' properties, capabilities, and limits are well-understood in a
huge range of applications. Using the same framework for species
interactions would thus allow ecologists to tap into an enormous set of
existing discoveries and techniques for dealing with indirect effects,
stability, and alternative stable states. Numerous other extensions are
possible: for example, the states of the interaction network can be
modeled as a function of the local abiotic environment (Lee and Hastie
2012), which would provide a rigorous and straightforward approach to
the difficult and important task of incorporating whole networks of
biotic interactions into species distribution models (Kissling et al.
2012, Pollock et al. 2014), leading to a better understanding of the
interplay between biotic and abiotic effects on community structure.
There are even methods (Whittam and Siegel-Causey 1981, Tjelmeland and
Besag 1998) that would allow one species to affect the sign or strength
of the relationship between two other species, tipping the balance
between facilitation and exploitation (Bruno et al. 2003).

Finally, the results presented here have important implications for
ecologists' continued use of null models for studying species
interactions. Null and neutral models can be useful for clarifying our
thinking about the numerical consequences of species' richness and
abundance patterns (Harris et al. 2011, Xiao et al. 2015), but
deviations from a particular null model must be interpreted with care
(Roughgarden 1983). Even in small networks with three species, it may
simply not be possible to implicate individual species pairs or specific
ecological processes like competition by rejecting a general-purpose
null (Gotelli and Ulrich 2009), especially when the test statistic is
effectively just a correlation coefficient (Figure 3B). Simultaneous
estimation of multiple ecological parameters seems like a much more
promising approach: to the extent that the models' relative performance
on real data sets is similar to the range of results shown in Figure 3A,
scientists in this field could often double their explanatory power by
switching from null models to Markov networks (or increase it
substantially with linear or generalized linear approximations).
Regardless of the methods ecologists ultimately choose, controlling for
indirect effects could clearly improve our understanding of species'
direct effects on one another and on community structure.

@@

@@

Xiao, X., D. J. McGlinn, and E. P. White. 2015. A strong test of the
Maximum Entropy Theory of Ecology. The American Naturalist 185:E70--E80.

\setlength{\parskip}{3pt} \setlength{\parindent}{0em}
\setlength{\leftskip}{0em}

\textbf{Figure 3.} \textbf{A.} Proportion of variance in interaction
coefficients explained by each method versus number of sampled
locations. \textbf{B.} The \(Z\)-scores produced by the null model
(``Pairs'') for each pair of species can be predicted using the
correlation between the presence-absence vectors of those same species
and from the number of sites on the landscape.

\end{flushleft}
\end{spacing}
